---
title: "Solutions for 10. Confounding variables"
subtitle: "Notes and in-class exercises"
sidebar: false
---




# Exercises

## Exercise 1: Review

The `peaks` data includes information on hiking trails in the 46 "high peaks" in the Adirondack mountains of northern New York state:


```{r message=FALSE, warning=FALSE}
# Load useful packages and data
library(readr)
library(ggplot2)
library(dplyr)
peaks <- read_csv("https://mac-stat.github.io/data/high_peaks.csv") %>%
    mutate(ascent = ascent / 1000)

# Check it out 
head(peaks)
```


```{r}
peaks_model <- lm(time ~ length + ascent + rating, data = peaks)
coef(summary(peaks_model))
```

- `length` coefficient:
    - Among hikes with the same vertical ascent and challenge rating, each additional mile of the hike is associated with a 0.46 hour increase in completion time on average.
    - Holding vertical ascent and challenge rating constant (fixed), each additional mile of the hike is associated with a 0.46 hour increase in completion time on average.

- `ratingeasy` coefficient:
    - Among hikes with the same length and vertical ascent, the average completion time of easy hikes is 3.2 hours less than that of difficult hikes (reference category).
    - Holding constant hike length and vertical ascent, the average completion time of easy hikes is 3.2 hours less than that of difficult hikes.




## Exercise 2: Confounders

```{r}
# Import data
cps <- read_csv("https://mac-stat.github.io/data/cps_2018.csv") %>% 
    filter(age >= 18, age <= 34) %>% 
    filter(wage < 250000)

# Check it out
head(cps)
```

age, education, job industry, ...

`marital` vs `wage`:

```{r}
cps %>%
  ggplot(aes(x=marital, y=wage))+
  geom_boxplot()+
  theme_classic()
```

`age` vs `marital`:

```{r}
cps %>%
  ggplot(aes(x=age, y=marital))+
  geom_boxplot()+
  theme_classic()
```

`age` vs `wage`:

```{r}
cps %>%
  ggplot(aes(x=age, y=wage))+
  geom_point()+
  geom_smooth(method="lm", se=F)+
  theme_classic()
```


## Exercise 3: Controlling for confounders

a. create 2 separate groups that are as similar as possible with respect to these variables. give the groups different remedies.

b. we'd have to get 2 groups that are similar with respect to age, and assign 1 group to get married and 1 group to be single. that would be weird (and unethical).



## Exercise 4: Age

```{r}
# Construct the model
wage_model_2 <- lm(wage ~ marital + age, cps)
coef(summary(wage_model_2))
```

a. .

```{r}
cps %>%
ggplot(aes(y = wage, x = age, color = marital)) +
    geom_point(size = 0.1, alpha = 0.5) +
    geom_line(aes(y = wage_model_2$fitted.values), size = 0.5)
```    

bonus! adding in the `geom_smooth` layer:

```{r}
cps %>%
ggplot(aes(y = wage, x = age, color = marital)) +
  geom_point(size = 0.1, alpha = 0.5) +
  geom_line(aes(y = wage_model_2$fitted.values), size = 0.5)+
  geom_smooth(method="lm", se=F, linetype="dashed")
```


b. -$7500


c. 
    - When controlling for ("holding constant") age, single workers make $7500 less than married workers on average.
    - Among workers of the same age, single workers make $7500 less than married workers on average.
    

    
    



## Exercise 5: More confounders

```{r}
wage_model_3 <- lm(wage ~ marital + age + education, cps)
coef(summary(wage_model_3))
```

a. 2 planes: There are 2 quantitative predictors which form the dimensions of the plane. The marital status categorical predictor creates 2 planes.

b. The `maritalsingle` coefficient is of main interest:
    - Among workers of the same age and years of education, single workers earn $6478 less than married workers.

c.
    - `age` coefficient: Among workers of the same marital status and years of education, each additional year of age is associated with a $1677 increase in salary on average.
    - `education` coefficient: Among workers of the same marital status and age, each additional year of education is associated with a $4285 increase in salary on average.





## Exercise 6: Even more

```{r}
wage_model_4 <- lm(wage ~ marital + age + education + industry, cps)
coef(summary(wage_model_4))
```

a. Among workers of the same job industry, education, and age, single workers make $5893 less than a married worker on average.

b. highest = construction (because it has the highest positive coefficient), lowest = service (because it has the most negative coefficient)





## Exercise 7: Biggest model yet

```{r}
wage_model_5 <- lm(wage ~ marital + age + education + industry + hours + health, cps)
coef(summary(wage_model_5))
```





## Exercise 8: Reflection

a. These confounders explained away more and more of the wage gap between single and married workers.
b. Answers will vary. A potential factor that we haven't considered is a worker's role within a given industry.













## Exercise 9: A new (extreme) example

clarity  description
-------- ------------------------------------------------
IF       flawless (no internal imperfections) 
VVS1     very very slightly imperfect
VVS2     " "
VS1      very slightly imperfect
VS2      " "
SI1      slightly imperfect
SI2      " "
I1       imperfect

```{r}
diamond_model_1 <- lm(price ~ clarity, data = diamonds)

# Get a model summary
coef(summary(diamond_model_1))
```

a. highest = SI2, lowest = VVS1

b. will vary.
    
    





## Exercise 10: Size

```{r}
diamond_model_2 <- lm(price ~ clarity + carat, data = diamonds)

# Get a model summary
coef(summary(diamond_model_2))

# Plot the model
diamonds %>% 
    ggplot(aes(y = price, x = carat, color = clarity)) + 
    geom_line(aes(y = diamond_model_2$fitted.values))
```

highest = IF, lowest = I1 (reference category)

This is what we would have expected!








## Exercise 11: Simpson's Paradox

The bigger the diamond the bigger the price:

```{r}
diamonds %>% 
    ggplot(aes(y = price, x = carat)) + 
    geom_point()
```


BUT the bigger the diamond, the more flawed it tends to be:

```{r}
diamonds %>% 
    ggplot(aes(y = carat, x = clarity)) + 
    geom_boxplot()
```


Thus flawed diamonds looked more expensive, but only because they also tend to be bigger (and size is a bigger driver of price).




## Exercise 12: Final conclusion

Flawless diamonds are more expensive.
    

