# Introduction to multiple regression


```{r include = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, 
  warning = FALSE,
  message = FALSE,
  error = TRUE)
```


<!-- create student notes from activity in core-->

{{< include _create_student_notes.qmd >}}

```{r}
#| echo: false
#| eval: true
make_student_notes('Introduction to multiple regression', '_L08-mlr-intro.qmd')
```

<!-- pull activity from core -->

```{r knitr-opts}
#| echo: false
#| cache: false
knitr::opts_chunk$set(eval = FALSE, echo = TRUE)
```

{{< include core/_L08-mlr-intro.qmd >}}



\
\
\
\



# Solutions

```{r eval = TRUE, echo = FALSE}
# Load packages
library(readr)
library(ggplot2)
library(dplyr)

# Load data
library(palmerpenguins)
data(penguins)
penguins <- penguins %>% 
  filter(species != "Adelie", bill_length_mm < 57)

# Import data
cps <- read_csv("https://mac-stat.github.io/data/cps_2018.csv") %>% 
  select(-education, -hours) %>% 
  filter(age >= 18, age <= 34) %>% 
  filter(wage < 250000)
```

## Exercise 1: Visualizing the relationship

a. no wrong answer

b. There are multiple options!

```{r eval = TRUE}
penguins %>% 
  ggplot(aes(y = flipper_length_mm, x = bill_length_mm, color = species)) + 
  geom_point()

penguins %>% 
  ggplot(aes(y = flipper_length_mm, x = bill_length_mm, shape = species)) + 
  geom_point()
```










## Exercise 2: Visualizing the model

a. no wrong answer

b. 

```{r eval = TRUE}
penguins %>% 
  ggplot(aes(y = flipper_length_mm, x = bill_length_mm, color = species)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)
```




## Exercise 3: Intuition

a. Gentoo tend to have longer flippers.
b. Flipper length is positively associated with bill length.
c. No. the lines are parallel / have the same slopes.









## Exercise 4: Model formula


a.  `bill_length_mm + species`
b. E[flipper_length_mm | bill_length_mm, speciesGentoo] = 127.75 + 1.40 * bill_length_mm + 22.85 * speciesGentoo






## Exercise 5: Sub-model formulas

Chinstrap: flipper_length_mm = 127.75 + 1.40 bill_length_mm

Gentoo:    flipper_length_mm = (127.75 + 22.85) + 1.40 bill_length_mm = 150.6 + 1.40 bill_length_mm





## Exercise 6: coefficients -- physical interpretation

a. The intercept coefficient, 127.75, is the intercept of the line for **Chinstrap** penguins.

b. The `bill_length_mm` coefficient, 1.40, is the **slope** of both lines.

c. The `speciesGentoo` coefficient, 22.85, indicates that the **intercept** of the line for Gentoo is 22.85mm higher than the **intercept** of the line for Chinstrap. Similarly, since the lines are parallel, the line for Gentoo is 22.85mm higher than the line for Chinstrap at any `bill_length_mm`.
    
    



## Exercise 7: coefficients -- contextual interpretation

a. For Chinstrap penguins with 0mm bills (silly), we expect a flipper length of 127.75mm.

b. For both Chinstrap and Gentoo penguins, flipper lengths increase by 1.40mm on average for every additional mm in bill length.

c. At any `bill_length_mm`, we expect the a Gentoo penguin to have 22.85mm longer flippers than a Chinstrap, on average.





## Exercise 8: Prediction

```{r eval = TRUE}
# a
127.75 + 1.40*50 + 22.85*0

# b
127.75 + 1.40*50 + 22.85*1

# c
predict(penguin_mod,
        newdata = data.frame(bill_length_mm = 50, 
                             species = "Chinstrap"))
predict(penguin_mod,
        newdata = data.frame(bill_length_mm = 50, 
                             species = "Gentoo"))
```






## Exercise 9: R-squared

a. species
b. no wrong answer
c. It's higher than the R-squared when we use either predictor alone!

