# Multiple linear regression: exploring interaction


```{r include = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, 
  warning = FALSE,
  message = FALSE,
  error = TRUE)
```


<!-- create student notes from activity in core-->

{{< include _create_student_notes.qmd >}}

```{r}
#| echo: false
#| eval: true
make_student_notes('Multiple linear regression: exploring interaction', '_L11-mlr-interaction-explore.qmd')
```

<!-- pull activity from core -->

```{r knitr-opts}
#| echo: false
#| cache: false
knitr::opts_chunk$set(eval = FALSE, echo = TRUE)
```

{{< include core/_L11-mlr-interaction-explore.qmd >}}



\
\
\
\


# Solutions

```{r eval = TRUE, echo = FALSE}
# Load packages and data
library(readr)
library(ggplot2)
library(dplyr)
cps <- read_csv("https://mac-stat.github.io/data/cps_2018.csv")

# Get data on just the management and transportation sectors
cps_sub <- cps %>% 
    filter(industry %in% c("management", "transportation"))
```

## Exercise 1: Wages across all industries

The plot below illustrates the relationship between wage and education for all of the industries in our `cps` dataset.

```{r eval = TRUE}
# Plot
ggplot(cps, aes(y = wage, x = education, color = industry)) + 
    geom_smooth(method = "lm", se = FALSE)
```

a. The industry-specific lines all have different slopes.

b. ag (first in alphabetical order)

c. This is a challenge! Compare your prediction to what you see when fitting the model in part d.

d. 

```{r eval = TRUE}
# Fit an interaction model called wage_model
wage_model <- lm(wage ~ education*industry, data = cps)

# Display summarized model output
coef(summary(wage_model))
```

e. In the **management** industry, wages increase the *most* per year of education. The increase is 61.95396 + 8616.31984 = \$8678.274 per year. That is, every additional year of education is associated with an average increase of \$8678.27 in yearly wages in the management industry.

f. In the **ag**riculture industry, wages increase the *least* per year of education. The increase is \$61.95 per year.  That is, every additional year of education is associated with an average increase of \$61.95 in yearly wages in the ag industry.


## Exercise 2: Thinking beyond

If a variable `x` has an interaction with the `industry` variable in affecting wages, then the relationship between `x` and wages must be different by industry. We might suspect that this could be the case for `hours` worked per week. We can make a plot to verify that this is actually the case:

```{r eval = TRUE}
ggplot(cps, aes(y = wage, x = hours, color = industry)) + 
    geom_smooth(method = "lm", se = FALSE)
```


