# Hypothesis testing: discovery



```{r include = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, 
  warning = FALSE,
  message = FALSE,
  error = TRUE)
```


<!-- create student notes from activity in core-->

{{< include _create_student_notes.qmd >}}

```{r}
#| echo: false
#| eval: true
make_student_notes('Hypothesis testing: discovery', '_L22-hypothesis-testing-discovery.qmd')
```

<!-- pull activity from core -->

```{r knitr-opts}
#| echo: false
#| cache: false
knitr::opts_chunk$set(eval = FALSE, echo = TRUE)
```

{{< include core/_L22-hypothesis-testing-discovery.qmd >}}





\
\
\
\




# Solutions



```{r eval = TRUE}
# Load the data & packages
library(tidyverse)
fish <- read_csv("https://mac-stat.github.io/data/Mercury.csv")
head(fish)
```


## Exercise 1: Review


### Part a

```{r eval = TRUE}
fish_mod_1 <- lm(Concen ~ Length, data = fish)
summary(fish_mod_1)
```

We estimate that for every 1cm increase in length, the expected mercury concentration increases by 0.058ppm.

### Part b

$0.058 \pm 2*0.005 = (0.058 - 0.010, 0.058 + 0.010) = (0.048, 0.068)$

```{r eval = TRUE}
confint(fish_mod_1, level = 0.95)
```


### Part c

We're 95% confident that for every 1cm increase in length, the expected mercury concentration increases by somewhere between 0.048ppm and 0.068ppm.


## Exercise 2: Test a hypothesis using the CI

### Part a 

The simple estimate $\hat{\beta}_1$ has error that we need to account for when making inferences about the population.
The CI accounts for this error!

### Part b

Yes, the interval doesn't include 0.
Thus 0 is not a plausible value for $\beta_1$.


## Exercise 3: Simulating data under the null value

### Part a

Intuition.

### Part b


```{r eval = TRUE}
# First check out the first 6 fish
fish %>% 
  select(Concen, Length) %>% 
  head()

# Run this chunk a few times!!
# Shuffle the Length values
fish %>%
  select(Concen, Length) %>% 
  mutate(Length = sample(Length, size = length(Length), replace = FALSE)) %>% 
  head()

# Shuffle the Length values
# Run this chunk a few times!!
# Then plot the resulting sample data and model
fish %>% 
  select(Concen, Length) %>% 
  mutate(Length = sample(Length, size = length(Length), replace = FALSE)) %>% 
  ggplot(aes(x = Length, y = Concen)) + 
  geom_point() +
  geom_smooth(method = "lm")
```



### Part c 

If there were truly no relationship between `Concen` and `Length`, we'd expect the sample model to have a slope near to (but not exactly) 0.

```{r eval = TRUE}
set.seed(1)
shuffled_models <- mosaic::do(500)*(
  fish %>% 
  select(Concen, Length) %>% 
  sample_n(size = length(Length), replace = TRUE) %>% 
  mutate(Length = sample(Length, size = length(Length), replace = FALSE)) %>%
  with(lm(Concen ~ Length))
)
head(shuffled_models)

fish %>% 
  ggplot(aes(x = Length, y = Concen)) + 
  geom_abline(data = shuffled_models, 
              aes(intercept = Intercept, slope = Length), 
              color = "gray", size = 0.25) + 
  geom_smooth(method = "lm", se = FALSE, size = 0) # Ignore this line. It's a clunky workaround
```

### Part d

We'd expect slopes to be Normally distributed around 0 (the null value).


```{r eval = TRUE}
shuffled_models %>% 
  ggplot(aes(x = Length)) + 
  geom_density()
```



## Exercise 4: Comparing our sample results to the null value (intuition)

### Part a

No! Its slope is much bigger than for the sample models simulated using the null value!

```{r eval = TRUE}
fish %>% 
  ggplot(aes(x = Length, y = Concen)) + 
  geom_abline(data = shuffled_models, 
              aes(intercept = Intercept, slope = Length), 
              color = "gray", size = 0.25) + 
  geom_smooth(method = "lm", se = FALSE)
```


### Part b

No! Our slope is much bigger than for the sample models simulated using the null value!

```{r eval = TRUE}
shuffled_models %>% 
  ggplot(aes(x = Length)) + 
  geom_density() + 
  geom_vline(xintercept = 0.05813, color = "blue")
```


### Part c 

Will vary.


## Exercise 5: CLT

The simulated and CLT / theory-informed sampling distributions are similar!

```{r eval = TRUE}
shuffled_models %>% 
  ggplot(aes(x = Length)) + 
  geom_density()

clt_plot <- data.frame(x = 0 + c(-4:4)*0.005) %>% 
  mutate(y = dnorm(x, sd = 0.005)) %>% 
  ggplot(aes(x = x)) +
  stat_function(fun = dnorm, args = list(mean = 0, sd = 0.005)) +
  geom_segment(aes(x = x, xend = x, y = 0, yend = y), linetype = "dashed") + 
  scale_x_continuous(breaks = c(-4:4)*0.005) + 
  xlab("")
  
clt_plot
```





## Exercise 6: Comparing our sample results to the null value (test statistic)

```{r eval = TRUE}
clt_plot + 
  geom_vline(xintercept = 0.05813, color = "blue")
```

### Part a 


- How far is *our* sample slope estimate of $\hat{\beta}_1 = 0.05813$ from the null value of 0? 0.05813 :)

- In the context of this particular analysis, is that a big or a small number? How can you tell? This seems pretty big relative to the standard error.


### Part b 

Our sample slope of 0.058 falls more than 11 standard errors above the null value of 0.
That's far!!!

```{r}
(0.058127 - 0) / 0.005228
```

### Part c

In the `t value` column:

```{r}
coef(summary(fish_mod_1))
```

### Part d

- Our sample data is *not* consistent with the null value of $\beta_1 = 0$, i.e. the idea that there's no significant association between mercury concentration and length.





## Exercise 7: Comparing our sample results to the null value (p-value)


```{r eval = TRUE}
clt_plot + 
  geom_vline(xintercept = 0.05813, color = "blue")
```


### Part a

Our estimate is more than 3 s.e. away from 0. Since 99.7% of estimates fall within 3 s.e. the probability of this happening is...

- less than 0.003 (1 - 0.997)


### Part b

< 2e-16 (which is very very close to 0)

```{r eval = TRUE}
coef(summary(fish_mod_1))
```

### Part c


- It's very unlikely that we’d have observed such a steep increase in `Concen` with `Length` among our sample fish "by chance", i.e. if in fact there were no relationship between mercury concentration and length in the broader fish population.




## Exercise 8: River test Part I


### Part a 

$\beta_1$, the RiverWacamaw coefficient

The null value is $\beta_1 = 0$

### Part b

$\hat{\beta}_1 = 0.142$ with a standard error of 0.089:

```{r eval = TRUE}
fish_mod_2 <- lm(Concen ~ River + Length, data = fish)

coef(summary(fish_mod_2))
```

### Part c

```{r eval = TRUE}
# Put OUR sample estimate here
# Round to 3 digits
est <- 0.142

# Put the corresponding standard error here
# Round to 3 digits
se <- 0.089

data.frame(x = 0 + c(-4:4)*se) %>% 
  mutate(y = dnorm(x, sd = se)) %>% 
  ggplot(aes(x = x)) +
  stat_function(fun = dnorm, args = list(mean = 0, sd = se)) +
  geom_segment(aes(x = x, xend = x, y = 0, yend = y), linetype = "dashed") + 
  scale_x_continuous(breaks = c(-4:4)*se) + 
  geom_vline(xintercept = est, color = "blue")
  xlab("")
```


## Exercise 9: River test Part II

### Part a

Yes! Our estimate is relatively close to 0 on this scale (within 2 s.e.).


### Part b

Our estimate of $\beta_1$, 0.142, falls only 1.59 s.e. from 0:

```{r eval = TRUE}
(0.142 - 0) / 0.089
coef(summary(fish_mod_2))
```


### Part c

Our estimate is somewhere between 1 and 2 s.e. from 0:

- between 0.05 and 0.32

An exact p-value is 0.114:

```{r}
coef(summary(fish_mod_2))
```


### Part d

It is. It is not very far from 0 (when considering s.e.) and would not be unlikely to observe if the null value were true.


### Part e

When controlling for length...

- we do not have sufficient evidence to conclude that there's a statistically significant difference in mercury concentration by river.



## Exercise 10: River CI

### Part a

Rough version:

$0.142 \pm 2*0.089 = (-0.036, 0.320)$

Exact version:

```{r}
confint(fish_mod_2)
```

### Part b

When controlling for length...

- we do not have sufficient evidence to conclude that there's a statistically significant difference in mercury concentration by river.

